Research log

### 2024-01-07 docs/paper.Rmd
Ian's comments

---
title             : "Pretense reveals a capacity for online simulation of counterfactual belief states"
shorttitle        : "Pretense reveals a capacity for online simulation of counterfactual belief states"

author: 
  - name          : "Matan Mazor"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Malet Street, London WC1E 7HX"
    email         : "mtnmzor@gmail.com"
  #   role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
  #     - Conceptualization
  #     - Writing - Original Draft Preparation
  #     - Writing - Review & Editing

  - name          : "Chaz Firestone"
    affiliation   : "3"

  - name          : "Ian Phillips"
    affiliation   : "3"
    # role:
    #   - Writing - Review & Editing

affiliation:
  - id            : "1"
    institution   : "Birkbeck, University of London"
  - id            : "2"
    institution   : "Wellcome Centre for Human Neuroimaging, UCL"
  - id            : "3"
    institution   : "Johns Hopkins University"

abstract: |
  What we do depends on what we know. But sometimes we try to decouple our behavior from our knowledge so as to appear not to know what we really do know. Such pretense behavior requires understanding how we would behave with different knowledge — and so provides a window into counterfactual self-simulation. However, little research has characterized and evaluated pretense relative to non-pretense behavior. Here, in two large-scale, pre-registered experiments, we tested people’s ability to conceal knowledge in the context of information-seeking games (Battleship and Hangman). Participants played normal games, as well as ‘pretend’ games in which they had full access to the game state, but pretended they didn’t. Overall, we find a highly impressive capacity for pretense — so impressive that independent 'judge' participants were completely unable to discriminate between normal and ‘pretend’ games. Specifically, pretenders successfully reproduced subtle features of real players' decision-making and timing. Nonetheless, pretense was characterized by pervasive differences (undetected by judges but revealed by follow-up analyses): pretenders played less rationally than real players, and had an increased tendency to rely on heuristics and rules. Our findings thus reveal a powerful but imperfect ability to simulate decisions and behaviour based on a counterfactual belief state, and open up a new approach to studying self-simulation.

@@

keywords          : "pretense; metacognition; theory of mind"
wordcount         : "X"

@@

Pretense relies on an ability to simulate and mimic one's own behavior under a hypothetical belief state. For example, in order to successfully deceive your friends into thinking that you were surprised by the birthday party they threw for you, it is not sufficient that you are able to reason about their mental states ("I know that they are planning a surprise party, but they don't know that I know that.") --- you also need to convincingly simulate and mimic your hypothetical behavior had you not known about the party ("Where would I look first? What would I say? How long would it take me to recover from the surprise?"). This is not a trivial challenge: previous research demonstrates that access to knowledge distorts our beliefs about what we would have thought [@fischhoff1975] or perceived [@harley2004] without that knowledge. Indeed, based on previous research, it is possible that people are completely unable to mentally simulate their own behaviour under a counterfactual state of ignorance.

<!-- to what extent humans are capable of generating complex behaviour that is aligned with a counterfactual belief state, and specifically with a counterfactual state of ignorance. -->

To this end, we examine pretense in a game setting. Using an online version of the games Battleships and Hangman, participants played a 'non-pretend' (normal) version of the game, as well as a 'pretend' version where they were given secret information about the location of hidden ships/target word, but were instructed to behave as if they didn't have this information. Overall, we find a highly impressive capacity for pretense, with pretenders mirroring subtle features of real players' decision-making and timing. At the same time, pretense was characterized by pervasive differences, with pretenders consistently playing less rationally than normal players, sometimes rigidly following heuristics and rules. Whilst powerful, then, our capacity for self-simulation is imperfect, displaying tell-tale biases in our internal models of our own cognition and decision making.

@@

In two experiments, players played online versions of two information-seeking games: Battleship (N=`r E2.df$subj_id%>%unique()%>%length()`) and Hangman (N=`r E4.df$subj_id%>%unique()%>%length()`). These games typically start in a state of ignorance, where players' goal is to reveal an unknown world state (grid configuration in Battleship, a hidden word in Hangman) with as few steps (cell or letter selections) as possible. Critically, in addition to playing five games in their standard version, players also completed five pretend games in which the true world state was known to them from the start, but their task was to behave as if it was not (see Fig. \@ref(fig:design)). We measured their capacity to simulate a counterfactual knowledge state of ignorance by comparing patterns of decisions and decision times in pretend and non-pretend games. Our full pre-registered results are available [online](https://matanmazor.github.io/ignorance/docs/exp.-1-battleship.html).

```{r design, echo=FALSE, fig.cap="Experimental Design in Exp. 1 (left) and 2 (right). In non-pretend games, players revealed ships by guessing cells in a grid (A) or revealed a word by guessing letters (E). In pretend games, we marked ship locations with a cross (B) and revealed the target word from the start (F), but asked players to play as if they didn't have access to this information. In half games (C and G), players continued the game from a given state, with the content of some cells and the position of some letters already revealed. Lastly, players watched replays of the games of previous players and guessed which were pretend games (D and H).", out.width = '100%'}
knitr::include_graphics("figures/methods.png")
```

As a first gross measure of pretense quality, we compared the total number of guesses in pretend and non-pretend games. Among Battleship players, the number of cell selections was similar in pretend (mean = `r E2.median_clicks%>%pull(pretend)%>%mean()`) and non-pretend games (mean = `r E2.median_clicks%>%pull(nonpretend)%>%mean()`; $p=$ `r apa_print(E2.median_clicks$diff%>%t.test())$table$p.value`; Fig. \@ref(fig:decisions)A). In Hangman, pretenders tended to make about one additional letter guess on average than did non-pretenders, controlling for word length (pretend: `r E4.median_misses%>%pull(pretend)%>%mean()`; non-pretend: `r E4.median_misses%>%pull(nonpretend)%>%mean()`; $p$ `r apa_print(E4.median_misses%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$table$p.value`; Fig. \@ref(fig:decisions)B). Despite an overall bias in the number of guesses, pretend Hangman games provided evidence for a more impressive, item-specific alignment: pretenders were successful in making more letter guesses when attempting to reveal words that would have been harder to guess had they been playing for real (`r cor.test(E4.misses_per_word$pretend, E4.misses_per_word$nonpretend,method='spearman')%>%apa_print()%>%'$'(estimate)`; Fig. \@ref(fig:decisions)B). This correlation provides a first clue for a human capacity to act in accordance with a counterfactual knowledge state.

(ref:decisions-caption) Battleship and Hangman guesses in pretend and non-pretend games. A: median number of misses Battleship and Hangman games, in non-pretend (green) and pretend (purple) games. For reference, the expected number of misses is indicated by a reference line for a fully random agent, and for a "greedy" agent that maximizes the probability of a hit in each step. B: The median number of misses in Hangman for pretend and non-pretend games, as a function of the target word. C: in half-games, players continued the game from a half-finished state. Showing spatial guess distributions for pretend and non-pretend games, alongside their corresponding hit probability maps. D: Cell and letter selections were ranked according to their relative hit probability. Plotting the median rank per subject in pretend and non-pretend games, with reference lines for the expected rank probability for a random agent, and for a "greedy" agent that maximizes the probability of a hit in each step. E: same as panel D but discarding all guesses that resulted in a hit.

@@

To further examine the decisional processes behind this strong alignment, we compared the degree to which pretend and non-pretend guesses made sense within the context of the game. When playing Battleship and Hangman, it makes sense to guess cells or letters for which the probability of hitting a ship or revealing a letter is high [this "greedy" behavior is not strictly optimal, but approximates optimal behavior in most cases, @audinot2014optimal, Section 3.3]. In the non-pretend versions of both games, guesses were more rational according to this measure than expected by chance (Battleship: `r E2.mean_P_click_rank%>%spread(test_part,p_click_rank)%>%mutate(diff=nonpretend-random)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`, Hangman: `r E4.mean_P_click_rank%>%spread(test_part,p_click_rank)%>%mutate(diff=nonpretend-random)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`). Despite being less rational than non-pretend guesses, pretend guesses were also more rational than expected by chance (Battleship: `r E2.mean_P_click_rank%>%spread(test_part,p_click_rank)%>%mutate(diff=pretend-random)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`, Hangman: `r E4.mean_P_click_rank_wide%>%pull(np_vs_r)%>%t.test()%>%apa_print%>%'$'(statistic)`; Fig. \@ref(fig:decisions)D). Critically, the same was true when restricting the analysis to unsuccessful guesses (Battleship: `r E2.mean_P_click_rank_misses_only%>%spread(test_part,p_click_rank)%>%mutate(diff=pretend-random)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`, Hangman: `r E4.mean_P_click_rank_misses_wide%>%pull(p_vs_r)%>%t.test()%>%apa_print%>%'$'(statistic)`; Fig. \@ref(fig:decisions)E): that is, even when missing ships or letters, pretend guesses made sense given the limited information players pretended to have.

Good pretense is a function not only of the number and content of players' decisions, but also of their timing. Here too, pretend games showed the same qualitative patterns as non-pretend games. Like non-pretenders, pretenders were faster in their successful guesses (Difference in decision time between hits and misses in Battleship: $\Delta_{non-pretend}=$ `r E2.RT_by_hit_lags %>% filter(test_part=='nonpretend')%>%pull(hit_diff)%>%median()%>%round()` ms, $\Delta_{pretend}=$ `r E2.RT_by_hit_lags %>% filter(test_part=='pretend')%>%pull(hit_diff)%>%median()%>%round()` ms; Hangman: $\Delta_{non-pretend}=$ `r E4.RT_by_hit_lags %>% filter(test_part=='nonpretend')%>%pull(hit_diff)%>%median(na.rm=T)%>%round()` ms, $\Delta_{pretend}=$ `r E4.RT_by_hit_lags %>% filter(test_part=='pretend')%>%pull(hit_diff)%>%median(na.rm=T)%>%round()` ms) and slowed down immediately after a hit (Difference in decision time between guesses that followed hits versus misses in Battleship: $\Delta_{non-pretend}=$ `r E2.RT_by_hit_lags %>% filter(test_part=='nonpretend')%>%pull(lag1_diff)%>%median()%>%round()` ms, $\Delta_{pretend}=$ `r E2.RT_by_hit_lags %>% filter(test_part=='pretend')%>%pull(lag1_diff)%>%median()%>%round()` ms; Hangman: $\Delta_{non-pretend}=$ `r E4.RT_by_hit_lags %>% filter(test_part=='nonpretend')%>%pull(lag1_diff)%>%median(na.rm=T)%>%round()` ms, $\Delta_{pretend}=$ `r E4.RT_by_hit_lags %>% filter(test_part=='pretend')%>%pull(lag1_diff)%>%median(na.rm=T)%>%round()` ms; Fig. \@ref(fig:time)A). Furthermore, decision uncertainty, quantified as the Shannon entropy of the posterior distribution over cell or letter options, had a similar quadratic effect on decision time in both non-pretend and pretend versions of the games, with the slowest responses associated with mid-range levels of entropy (see Fig. \@ref(fig:time)B).

```{r time, echo=FALSE, message= F, warning=F, fig.cap="Patterns of reaction time in pretend and non-pretend guesses. A: median decision times in and following hits and misses. In both Battleship and Hangman, hits were faster on average than misses, but guesses following a hit were slower on average than those following a miss. This pattern was mimicked in pretend games. B: median decision times as a function of decision uncertainty, quantified as the entropy of the posterior over guess options. In both Hangman and Battleship, guesses were slowest for mid-range levels of entropy, and this pattern was mimicked in pretend games.", out.width = '100%'}
knitr::include_graphics("figures/time.png")
```

This capacity for simulating a state of ignorance was not perfect. First, despite showing the same qualitative effects, decision times in Battleship (but not Hangman) pretend games were systematically more pronounced relative to non-pretend games, as if participants were acting based on an exaggerated mental image of their own counterfactual behaviour. Furthermore, in their attempt to behave as if they didn't know the true state of the game, pretenders had an increased tendency to follow rigid heuristics and rules, ignoring useful information as a result. The within-participant sequential guess entropy, a measure of the inverse predictability of decision sequences across different games of the same player, was significantly reduced in pretend games (Fig. \@ref(fig:rigidity)A; see [online supplementary materials](https://matanmazor.github.io/ignorance/docs/exp.-1-battleship-exploratory-results.html#within-participant-click-entropy)). More specifically, Hangman pretenders were more likely to guess letters that are frequent in the English language (E, T, A, etc.) irrespective of the game state, compared to genuine players (Fig. \@ref(fig:rigidity)B).

This limitation on incorporating evidence into the decision-making process was especially evident in Hangman half-games, where players completed the game from a half-completed state. When asked to reveal the hidden fruit "\_A_A_A", `r round(100*E4.first_click_half_games %>% filter(test_part=='nonpretend_half_game' & word %in% c('banana', 'papaya'))%>%pull(bananahand)%>%mean())`% of the non-pretenders guessed one of the letters 'B' or 'N' (Fig. \@ref(fig:rigidity)C, yellow bars in left column). Among pretenders who knew that the hidden word was BANANA, this preference was reduced to `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word== 'banana')%>%pull(bananahand)%>%mean())`%. Importantly, half of the pretenders were given different information: they were told that the hidden word is the less prototypical fruit PAPAYA. Although good pretenders should simulate their behavior had they not known this information, only `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word =='papaya')%>%pull(bananahand)%>%mean())`% selected the letters 'B' or 'N', revealing that many pretenders were unable to predict that the first fruit that would come to their minds is BANANA, not PAPAYA (Fig. \@ref(fig:rigidity)C, yellow bars in right column). A similar pattern was observed for the prototypical body part word HA(ND) and its surprising counterpart HA(IR): when playing normally, `r round(100*E4.first_click_half_games %>% filter(test_part=='nonpretend_half_game' & word %in% c('hand', 'hair'))%>%pull(bananahand)%>%mean())`% of the players selected letters that are consistent with the prototypical option HAND. Pretenders for whom the word was HAND showed the same tendency (`r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word =='hand')%>%pull(bananahand)%>%mean())`%), in contrast to only `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word =='hair')%>%pull(bananahand)%>%mean())`% of the pretenders who saw the word HAIR (Fig. \@ref(fig:rigidity)C, blue bars).

```{r rigidity, echo=FALSE, message= F, warning=F, fig.cap="Limitations on flexible decision-making when pretending. A: sequential guess entropy, a measure of the (inverse) predictability of individual players' guesses as a function of click number and guess number. In both Battleship and Hangman, Sequential guess entropy increased with click number, and was overall lower in pretend games. B: letter frequency of Hangman guesses: the median rank frequency per participant in pretend and non-pretend games, with reference lines for the expected rank frequency for a random agent, and for a “greedy” agent that maximizes the probability of a hit in each step. C: first letter guesses in Hangman half games, as a function of pretense condition and target word. Letters appear in alphabetical order. letters that appear in the target word are marked in yellow (BANANA), red (PAPAYA), blue (HAND) and magenta (HAIR). For reference, the overall distribution of letter guesses across all games and conditions is given below. ", out.width = '100%'}
knitr::include_graphics("figures/rigidity.png")
```

Despite these limitations, pretenders were able to fool other players into believing they were playing for real. After performing pretend and non-pretend games, participants were presented with game replays of previous players and judged who was pretending and who was playing for real. In both games, pretense detection was at chance (Battleship: `r E2.judge_df%>%group_by(subj_id)%>%summarise(accuracy=sum(correct)*20)%>%pull(accuracy)%>%mean()%>%round()`%, Hangman: `r round(E4.judge_summary_df$accuracy%>%mean()*100)`%). This is in line with previous findings of near-chance accuracy in lie detection [@bond2006]. Moreover, we find no sign of a correlation between pretense quality (measured as players' ability to trick judges to think they were not pretending) and pretense detection ability (measured as proportion correct; Battleship: `r apa_print(cor.test(E2.performance_df$judge_rate,E2.performance_df$pretend_rate, method='spearman'))$estimate`, Hangman: `r apa_print(cor.test(E4.performance_df$pretend_rate,E4.performance_df$judge_rate,method='spearman'))$estimate`), indicating that pretense and pretense detection rely on at least partly different cognitive processes.

@@

Participants were first instructed that the experiment, based on the game Battleships, had three parts, and that they could accumulate 'points' that would later translate to a monetary bonus payment. They were then presented with a leaderboard of previous players, and given the rules of the game:

>>>>

---
title             : "Pretending not to know: Pretense behavior reveals the limits of self-simulation"
shorttitle        : "Pretending not to know"

author: 
  - name          : "Matan Mazor"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "All Souls College, High Street, Oxford OX1 4AL"
    email         : "matan.mazor@all-souls.ox.ac.uk"
  #   role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
  #     - Conceptualization
  #     - Writing - Original Draft Preparation
  #     - Writing - Review & Editing

  - name          : "Chaz Firestone"
    affiliation   : "2"

  - name          : "Ian Phillips"
    affiliation   : "2"
    # role:
    #   - Writing - Review & Editing

affiliation:
  - id            : "1"
    institution   : "University of Oxford"
  - id            : "2"
    institution   : "Johns Hopkins University"

abstract: |
  What we do depends on what we know. But sometimes we try to decouple our behavior from our knowledge so as to appear not to know what we really do know. Such pretense behavior requires understanding how we would behave with different knowledge, and so provides a window into counterfactual self-simulation. However, little research has characterized and evaluated pretense relative to non-pretense behavior. Here, in two large-scale, pre-registered experiments, we tested people’s ability to conceal knowledge in the context of information-seeking games (Battleship and Hangman). Participants played normal games (in which they began with no knowledge of the ships’ locations or the word’s identity), as well as ‘pretend’ games in which they were shown the game's full solution but were instructed to play *as though* they were ignorant. We find a remarkable capacity for pretense, with pretenders successfully reproducing broad patterns and subtle features of real players’ decision-making and timing — so much so that 'judge' paticipants failed to distinguish real players from pretenders. Nonetheless, pretense was characterized by pervasive differences, undetected by judges but revealed by follow-up analyses: pretenders played less rationally than real players, and had an increased tendency to rely on heuristics and rules. Our findings thus reveal a powerful but imperfect ability to simulate decisions and behavior based on a counterfactual belief state, and open up a new approach to studying self-simulation.

@@

keywords          : "pretense; metacognition; theory of mind"
wordcount         : "5986"

@@

Pretense relies on an ability to simulate and mimic one's own behavior under a counterfactual belief state. For example, in order to successfully deceive your friends into thinking that you were surprised by the birthday party they threw for you, it is not sufficient that you are able to reason about their mental states ("I know that they are planning a surprise party, but they don't know that I know that...") --- you also need to convincingly simulate and mimic your hypothetical behavior had you not known about the party ("Where would I look first had I not known? What would I say? How long would it take me to recover from the surprise?"). This is not a trivial challenge: previous research on "hindsight biases" suggests that knowledge about the actual state of the world can interfere with our ability to correctly judge what we would have believed [@fischhoff1975; @fischhoff1977; @wood1978; @roese2012] or perceived [@harley2004; @bernstein2007; @bernstein2012] without this knowledge. Such biases remain potent even when instructing participants to overcome them [@harley2004; @pohl1996]. Moreover, even if pretenders can correctly determine what they would have believed, they must further accurately simulate how they would think and behave in this different belief state.

The reliance of this kind of epistemic pretense on self-simulation makes it a promising tool for revealing the structure and content of people's internal models of their own minds. When directly asked, participants are able to provide relatively accurate descriptions of their own decision-making [@morris] and perception [@levin2008; @mazor2023]. Pretending not to know opens a new window into the structure and content of this metacognitive knowledge, with two important advantages. First, by not relying on explicit reports, pretense has the potential to reveal implicit self-knowledge -- that is, structured knowledge about the self that is not reportable. And second, data obtained from pretense experiments can be analyzed and modeled using the same tools employed by cognitive scientists to study non-pretense behavior, affording a direct and finer-grained comparison between pretend and genuine decision-making.

<!-- to what extent humans are capable of generating complex behavior that is aligned with a counterfactual belief state, and specifically with a counterfactual state of ignorance. -->

To this end, we examine pretense in a game setting. Using an online version of the games Battleship and Hangman (in which players seek to uncover the locations of enemy ships or the identity of a word), participants played a 'non-pretend' (normal) version of the game, as well as a 'pretend' version where they were given complete information about the hidden ships / target word but were instructed to behave as if they didn't have this information. Overall, we find a highly impressive capacity for pretending not to know, with pretenders mirroring broad patterns and subtle features of real players' decisions and decision times. However, we also find that pretense was characterized by pervasive biases and limitations, although these were undetected by human observers.

@@

In two experiments, players played online versions of two information-seeking games: Battleship (N=`r E2.df$subj_id%>%unique()%>%length()` players) and Hangman (N=`r E4.df$subj_id%>%unique()%>%length()` players). These games traditionally start in a state of ignorance, with a player's goal being to reveal an unknown world state (ship locations in Battleship, a hidden word in Hangman) in as few steps (cell or letter selections) as possible. Critically, in addition to playing five standard games, players in our experiments also completed five 'pretend' games in which the solution to the game was known to them from the start, but where their task was to behave as if it was not -- i.e. to play as though they did not have this information (see Fig. \@ref(fig:design)). We measured participants' capacity to simulate a counterfactual state of ignorance by comparing patterns of decisions and decision times in pretend and non-pretend games. Our full pre-registered results are available [online](https://matanmazor.github.io/ignorance/docs/exp.-1-battleship.html) together with the report-generating code. Readers are invited to try [demos of the experiments](https://matanmazor.github.io/ignorance/experiments/demos/pretend).

```{r design, echo=FALSE, fig.cap="Experimental Design in Exp. 1 (left) and 2 (right). In non-pretend games, players revealed ships by guessing cells in a grid (A) or revealed a word by guessing letters (E). In pretend games, we marked ship locations with a cross (B) and revealed the target word from the start (F), but asked players to play as if they didn't have this information. In half-games (C and G), players continued the game from a given state, with the content of some cells and the position of some letters already revealed. Lastly, players watched replays of the games of previous players and guessed which were pretend games (D and H).", out.width = '100%'}
knitr::include_graphics("figures/methods.png")
```

## Measuring pretense quality

As a first measure of pretense quality, we compared the total number of guesses in pretend and non-pretend games. Among Battleship players, the number of cell selections was similar in pretend (mean = `r E2.median_clicks%>%pull(pretend)%>%mean()%>%printnum()`) and non-pretend games (mean = `r E2.median_clicks%>%pull(nonpretend)%>%mean()%>%printnum()`; $p=$ `r apa_print(E2.median_clicks$diff%>%t.test())$table$p.value`; Fig. \@ref(fig:decisions)A). `r E2.lucky_pretenders%>%length` pretenders who immediately discovered all ships without making errors were excluded from all further analyses, in accordance with our pre-registered plan. With these subjects excluded, the number of cell selections remained very similar in pretend (mean = \`r `` E2.median_clicks_filtered%>%pull(pretend)%>%mean()%>%printnum()` ``) and non-pretend games (mean = \`r `` E2.median_clicks_filtered%>%pull(nonpretend)%>%mean()%>%printnum()` ``; $p=$`r apa_print(E2.median_clicks%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$table$p.value`; Fig. \@ref(fig:decisions)A). `r E2.lucky_pretenders%>%length`). In Hangman, pretenders tended to make about one additional letter guess on average than did non-pretenders, controlling for word length (pretend: `r E4.median_misses%>%pull(pretend)%>%mean()%>%printnum()` misses; non-pretend: `r E4.median_misses%>%pull(nonpretend)%>%mean()%>%printnum()` misses; $p$ `r apa_print(E4.median_misses%>%mutate(diff=pretend-nonpretend)%>%pull(diff)%>%t.test())$table$p.value`; Fig. \@ref(fig:decisions)B). Despite an overall bias in the number of guesses, pretend Hangman games showed an impressive, item-specific alignment: pretenders were successful in making more letter guesses when attempting to reveal words that would have been harder to guess had they been playing for real (`r cor.test(E4.misses_per_word$pretend, E4.misses_per_word$nonpretend,method='spearman')%>%apa_print()%>%'$'(estimate)`; Fig. \@ref(fig:decisions)B). This strong correlation provides evidence for a human capacity to act in accordance with a counterfactual knowledge state.

(ref:decisions-caption) Battleship and Hangman guesses in pretend and non-pretend games. A: median number of misses in Battleship and Hangman games, in non-pretend (green) and pretend (purple) games. For reference, the expected number of misses is indicated by a reference line for a fully random agent, and for a "greedy" agent that maximizes the probability of a hit in each step. B: The median number of misses in Hangman for pretend and non-pretend games, as a function of the target word. C: Spatial guess distributions for pretend and non-pretend half-games (where players continued the game from a half-finished state) alongside their corresponding hit probability maps. D: Cell and letter selections were ranked according to their relative hit probability. Plotting the median rank per subject in pretend and non-pretend games, with reference lines for the expected rank probability for a random agent, and for a "greedy" agent that maximizes the probability of a hit in each step. E: same as panel D but discarding all guesses that resulted in a hit.

@@

To further examine the decisional processes behind this strong alignment, we compared the degree to which pretend and non-pretend guesses made sense within the context of the game. When playing Battleship and Hangman, it makes sense to guess cells or letters for which the probability of hitting a ship or revealing a letter is high [this "greedy" behavior is not strictly optimal, but approximates optimal behavior in most cases, @audinot2014optimal, Section 3.3]. In the non-pretend versions of both games, guesses were more rational according to this measure than expected by chance (Battleship: `r E2.mean_P_click_rank%>%spread(test_part,p_click_rank)%>%mutate(diff=-nonpretend+random)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`, Hangman: `r E4.mean_P_click_rank%>%spread(test_part,p_click_rank)%>%mutate(diff=-nonpretend+random)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`). Despite being less rational than non-pretend guesses, pretend guesses were also more rational than expected by chance (Battleship: `r E2.mean_P_click_rank%>%spread(test_part,p_click_rank)%>%mutate(diff=-pretend+random)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`, Hangman: `r t.test(-E4.mean_P_click_rank_wide%>%pull(np_vs_r))%>%apa_print%>%'$'(statistic)`; Fig. \@ref(fig:decisions)D). Critically, the same was true when restricting the analysis to unsuccessful guesses (Battleship: `r E2.mean_P_click_rank_misses_only%>%spread(test_part,p_click_rank)%>%mutate(diff=-pretend+random)%>%pull(diff)%>%t.test()%>%apa_print()%>%'$'(statistic)`, Hangman: `r t.test(-E4.mean_P_click_rank_misses_wide%>%pull(p_vs_r))%>%apa_print%>%'$'(statistic)`; Fig. \@ref(fig:decisions)E): that is, even when incorrectly guessing a ship's location or a letter's identity, pretend guesses made sense given the limited information players pretended to have.

Good pretense is a function not only of the number and content of players' decisions, but also of their timing. Here too, pretend games showed the same qualitative patterns as non-pretend games. Like non-pretenders, pretenders were faster in their successful guesses (Difference in decision time between hits and misses in Battleship: $\Delta_{non-pretend}=$ `r E2.RT_by_hit_lags %>% filter(test_part=='nonpretend')%>%pull(hit_diff)%>%median()%>%round()` ms, $\Delta_{pretend}=$ `r E2.RT_by_hit_lags %>% filter(test_part=='pretend')%>%pull(hit_diff)%>%median()%>%round()` ms; Hangman: $\Delta_{non-pretend}=$ `r E4.RT_by_hit_lags %>% filter(test_part=='nonpretend')%>%pull(hit_diff)%>%median(na.rm=T)%>%round()` ms, $\Delta_{pretend}=$ `r E4.RT_by_hit_lags %>% filter(test_part=='pretend')%>%pull(hit_diff)%>%median(na.rm=T)%>%round()` ms) and slowed down immediately after a hit (Difference in decision time between guesses that followed hits versus misses in Battleship: $\Delta_{non-pretend}=$ `r E2.RT_by_hit_lags %>% filter(test_part=='nonpretend')%>%pull(lag1_diff)%>%median()%>%round()` ms, $\Delta_{pretend}=$ `r E2.RT_by_hit_lags %>% filter(test_part=='pretend')%>%pull(lag1_diff)%>%median()%>%round()` ms; Hangman: $\Delta_{non-pretend}=$ `r E4.RT_by_hit_lags %>% filter(test_part=='nonpretend')%>%pull(lag1_diff)%>%median(na.rm=T)%>%round()` ms, $\Delta_{pretend}=$ `r E4.RT_by_hit_lags %>% filter(test_part=='pretend')%>%pull(lag1_diff)%>%median(na.rm=T)%>%round()` ms; Fig. \@ref(fig:time)A). Furthermore, decision uncertainty, quantified as the Shannon entropy of the posterior distribution over cell or letter options, had a similar quadratic effect on decision time in both non-pretend and pretend versions of the games, with the slowest responses associated with mid-range levels of entropy (see Fig. \@ref(fig:time)B). In other words, despite knowing the game's solution with full certainty, pretenders successfully feigned subtle qualitative effects of counterfactual uncertainty on their decision times.

```{r time, echo=FALSE, message= F, warning=F, fig.cap="Patterns of decision time in pretend and non-pretend games. A: median decision times forhits and misses, as well as the decisions following them. In both Battleship and Hangman, hits were faster on average than misses, but guesses following a hit were slower on average than those following a miss. This pattern was mimicked in pretend games. B: median decision times as a function of decision uncertainty, quantified as the entropy of the posterior over guess options. In both Hangman and Battleship, guesses were slowest for mid-range levels of entropy, and this pattern was mimicked in pretend games.", out.width = '100%'}
knitr::include_graphics("figures/time.png")
```

## Pretense limitations

Though impressive, the capacity for simulating a state of ignorance was not perfect. First, despite showing the same qualitative effects, decision time patterns in Battleship pretend games (but not Hangman pretend games) were systematically more pronounced relative to non-pretend games, as if participants were acting based on an exaggerated mental model of their own counterfactual behavior. Furthermore, the within-participant sequential guess entropy, a measure of the inverse predictability of decision sequences across different games of the same player, was significantly reduced in pretend games (Fig. \@ref(fig:rigidity)A; see [online supplementary materials](https://matanmazor.github.io/ignorance/docs/exp.-1-battleship-exploratory-results.html#within-participant-click-entropy)). More specifically, Hangman pretenders were more likely to guess letters that appear frequently in English words (E, T, A, etc.) irrespective of the game state, compared to genuine players (Fig. \@ref(fig:rigidity)B). This suggests that in their attempt to behave as if they didn't know the true state of the game, pretenders had an increased tendency to follow rigid heuristics and rules, ignoring useful information as a result.

This limitation on incorporating evidence into the (simulated) decision-making process was especially evident in Hangman half-games, where players completed the game from a half-completed state. When asked to reveal the hidden fruit "\_A_A_A", `r round(100*E4.first_click_half_games %>% filter(test_part=='nonpretend_half_game' & word %in% c('banana', 'papaya'))%>%pull(bananahand)%>%mean())`% of the non-pretenders guessed one of the letters 'B' or 'N' (Fig. \@ref(fig:rigidity)C, yellow bars in left column). Among pretenders who knew that the hidden word was BANANA, this preference was reduced to `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word== 'banana')%>%pull(bananahand)%>%mean())`%. Importantly, half of the pretenders were given different information: they were told that the hidden word was the less prototypical fruit PAPAYA. Although good pretenders should simulate their behavior had they not known this information, only `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word =='papaya')%>%pull(bananahand)%>%mean())`% selected the letters 'B' or 'N', revealing that many pretenders were unable to predict that the first fruit that would have come to their minds was BANANA, not PAPAYA (Fig. \@ref(fig:rigidity)C, yellow bars in right column). A similar pattern was observed for the prototypical body part word HA(ND) and its surprising counterpart HA(IR): when playing normally, `r round(100*E4.first_click_half_games %>% filter(test_part=='nonpretend_half_game' & word %in% c('hand', 'hair'))%>%pull(bananahand)%>%mean())`% of the players selected letters that are consistent with the prototypical option HAND. This figure was `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word =='hand')%>%pull(bananahand)%>%mean())`% among pretenders for whom the target word was HAND, in contrast to only `r round(100*E4.first_click_half_games %>% filter(test_part=='pretend_half_game' & word =='hair')%>%pull(bananahand)%>%mean())`% among pretenders for whom the target word was HAIR (Fig. \@ref(fig:rigidity)C, blue bars).

```{r rigidity, echo=FALSE, message= F, warning=F, fig.cap="Limitations on flexible decision-making when pretending. A: sequential guess entropy, a measure of the (inverse) predictability of individual players' guesses as a function of click number and guess number. In both Battleship and Hangman, sequential guess entropy increased with click number, and was overall lower in pretend games. B: letter frequency of Hangman guesses: the median rank frequency per participant in pretend and non-pretend games, with reference lines for the expected rank frequency for a random agent, and for a “greedy” agent that maximizes the probability of a hit in each step. C: first letter guesses in Hangman half-games, as a function of pretense condition and target word. Letters appear in alphabetical order. letters that appear in the target word are marked in yellow (BANANA), red (PAPAYA), blue (HAND) and magenta (HAIR). For reference, the overall distribution of letter guesses across all games and conditions is given below. ", out.width = '100%'}
knitr::include_graphics("figures/rigidity.png")
```

## Failure to detect pretending

Despite these limitations, pretenders' behavior resembles real players' behaviour closely enough that they were able to fool other participants into believing they were playing for real. After performing pretend and non-pretend games, participants were presented with game replays of previous players and took on the new role of being a 'judge' who had to determine who was pretending and who was playing for real. In both games, pretense detection was at chance (Battleship: `r E2.judge_df%>%group_by(subj_id)%>%summarise(accuracy=sum(correct)*20)%>%pull(accuracy)%>%mean()%>%round()`%, Hangman: `r round(E4.judge_summary_df$accuracy%>%mean()*100)`%). This is in line with previous findings of near-chance accuracy in lie detection [@bond2006]. Moreover, we find no sign of a correlation between pretense quality (measured as players' ability to trick judges into thinking they were not pretending) and pretense detection ability (measured as proportion correct; Battleship: `r apa_print(cor.test(E2.performance_df$judge_rate,E2.performance_df$pretend_rate, method='spearman'))$estimate`, Hangman: `r apa_print(cor.test(E4.performance_df$pretend_rate,E4.performance_df$judge_rate,method='spearman'))$estimate`), indicating that pretense and pretense detection rely on at least partly different cognitive processes.

@@

Participants were first instructed that the experiment, based on the game Battleship, had three parts, and that they could accumulate 'points' that would later translate to a monetary bonus payment. They were then presented with a leaderboard of previous players, and given the rules of the game:

> *"In the game Battleship, your task is to sink all ships located in a grid with as few clicks as possible. What makes the game difficult is that you can't see the ships; all you can see is a grid of squares, and you have to guess where the ships are. To sink a ship, you need to click on all of the squares it is located in. If you hit part of a ship, the square will turn red. If there is no ship in the square, it will turn blue.*"
